Ricci-Ollivier Curvature of the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-03-31 curvature.tex
Spelling pass.

\begin{figure}
	\hspace*{\stretch{1}}
	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
	\hspace*{\stretch{2}}
	\subfigure[7 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter7}}
	\hspace*{\stretch{1}}
    \caption{Scatterplot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph. Color displays the average degree of $T_1$ and $T_2$. Distance values randomly perturbed (``jittered'') a small amount to avoid overplotting.}
	\label{fig:rspr-scatter}
\end{figure}

@@

\begin{lemma}
	\label{lem:select_random_neighbor}
	Given a tree $T$, an rSPR neighbor of $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
\end{lemma}
\begin{proof}
	We apply the above procedure.
	We use a standard nodes-and-pointers representation of the trees, which can be constructed in $\OhOf{n}$-time from a newick string representation and uses linear space in $n$.
	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
	We precompute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
	Both of these traversals require $\OhOf{n}$-time.
	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding rSPR destination.
	Finally, we require linear time to apply the chosen rSPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
	Thus, the for loop requires linear time.
	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
	Therefore, the procedure uses linear time and space and selects an rSPR neighbor of $T$ uniformly at random.
\end{proof}

@@

We next bound the difference between the coarse and asymptotic curvatures.
Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that remains at a given tree with probability $1-p$ and transitions with probability $p$.
For the lazy uniform random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
The asympototic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
As we now prove, these two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$.

@@

	In the case that $\dspr{T,S}=1$, the affected mass must be moved through only at most one additonal tree, as $T$ and $S$ are adjacent.
	We thus obtain the lower bound of $\curvature{T,S} - \frac{1}{6}$ in this case.

>>>>

\begin{figure}
	\hspace*{\stretch{1}}
	\subfigure[6 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter6}}
	\hspace*{\stretch{2}}
	\subfigure[7 taxa]{\includegraphics[width=3in]{figs/rspr-upmh-scatter7}}
	\hspace*{\stretch{1}}
    \caption{Scatter plot of $\curvature{\MH; T_1,T_2}$ values versus $\dspr{T_1,T_2}$ for the rSPR graph. Color displays the average degree of $T_1$ and $T_2$. Distance values randomly perturbed (``jittered'') a small amount to avoid superimposed points.}
	\label{fig:rspr-scatter}
\end{figure}

@@

\begin{lemma}
	\label{lem:select_random_neighbor}
	Given a tree $T$, an rSPR neighbor of $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
\end{lemma}
\begin{proof}
	We apply the above procedure.
	We use a standard nodes-and-pointers representation of the trees, which can be constructed in $\OhOf{n}$-time from a Newick string representation and uses linear space in $n$.
	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
	We pre-compute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
	Both of these traversals require $\OhOf{n}$-time.
	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding rSPR destination.
	Finally, we require linear time to apply the chosen rSPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
	Thus, the for loop requires linear time.
	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
	Therefore, the procedure uses linear time and space and selects an rSPR neighbor of $T$ uniformly at random.
\end{proof}

@@

We next bound the difference between the coarse and asymptotic curvatures.
Recall that $\curvature[p]{T,S}$ is the coarse Ricci-Ollivier curvature between trees $T$ and $S$ with respect to the lazy walk that remains at a given tree with probability $1-p$ and transitions with probability $p$.
For the lazy uniform random walk, $m_T$ is now $T \cup N(T)$, with each neighbor assigned mass $\frac{p}{\degree{T}}$ and $T$ assigned the remaining $1 - p$ mass.
The asymptotic Ricci-Ollivier curvature $\ric{T,S}$ is $\lim_{p \rightarrow 0} \curvature[p]{T,S} / p$.
As we now prove, these two notions of curvature differ only by a small factor inversely proportional to the maximum degree of $T$ and $S$.

@@

	In the case that $\dspr{T,S}=1$, the affected mass must be moved through only at most one additional tree, as $T$ and $S$ are adjacent.
	We thus obtain the lower bound of $\curvature{T,S} - \frac{1}{6}$ in this case.
